$Z_3$ topological order in the quantum dimer-pentamer model

### 2017-01-06 qdpm_paper.tex
Worked on text

If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex, within the context of the original GQDM. One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function; if the free energy of defects at large distances remains finite, electric charges are deconfined in the GQDM ground state.

@@

We now consider relaxing the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{\rm{IGT}} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{\rm{IGT}} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{\rm{IGT}}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega: (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

We may understand the emergence of the $Z_2$ topologically ordered phase of IGTs from the QDM and QFPLM from the following picture. The additional vertex configurations that are allowed in the IGTs relative to the models with a dimer number constraint (monomers and pentamers in the even sector and tetramers in the odd sector, see \tabref{tab:defects}) have charge U(1) charge $Q_v = \pm 2$. It is known that coupling a $U(1)$ gauge field to a dynamics charge $q$ matter field can result in an $Z_q$ gauge theory~\cite{Fradkin1979}; as in the IGT's we have added dynamic charge 2 defects to the underlying $U(1)$ symmetric model, and we may expect a $Z_2$ gauge theory to result. We note that pentamers are charge $\pm3$ objects in the QDM (see \tabref{tab:defects}); thus below we investigate whether introducing pentamers into the QDM leads to an effective $Z_3$ gauge theory.

@@

The Hamiltonian of the QDPM is defined by \Eqref{eq:HGQDM} including both plaquette flips as well as local dimer resonances that create and destroy pairs of pentamers, and allow single pentamers to translate on the same sub-lattice; these local dynamics involving pentamers are illustrated in  \tabref{tab:pentamer_moves}. Thus we may write $\HQDPM$ as the following:
\begin{equation}
\HQDPM \equiv \HQDM + H_{\rm{pen}}
\end{equation}
where $H_{\rm{pen}}$ only involves the pentamer resonances in \tabref{tab:pentamer_moves} and those related by symmetry:
\begin{align}
H_{\rm{pen}} \equiv \HGQDM \biggl( &\bigl\{ \left(c_a,\bar{c}_a\right),\left(c_b,\bar{c}_b\right),\left(c_c,\bar{c}_c\right) \bigr\},\biggr.\notag\\
\biggl. &\bigl\{ \left(t_a,v_a\right),\left(t_b,v_b\right),\left(t_c,v_c\right) \bigr\} \biggr)  \label{eq:Hpen}
\end{align}

\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
    $\nu$ & $c_\nu$ & $\bar{c}_\nu$ \\
	\hline
    $a$ & $\pentmvcreate$ &  $\pentmvanal$  \\
	\hline
    $b$ &$\penmvhrzleft$&$\penmvhrzright$ \\
	\hline
    $c$ &$\penmvdiagup$& $\penmvdiagdown$\\
	\hline
  \end{tabular}
  \caption{The pairs of local dimer \& pentamer configurations $c_\nu\leftrightarrow\bar{c}_\nu$ generate minimal local dynamics that create/annihilate pairs ($a$) of pentamers and translate single pentamers ($b$ \& $c$). Note that pentamers always stay on the same sublattice which is associated with their $U(1)$ charge. }
  \label{tab:pentamer_moves}
\end{table}

Because of the terms involving pentamers, $\HQDPM$ conserves $Q_v^{Z_3}$ rather than $Q_v$, and thus $\Phi_\Gamma$ does not commute with the Hamiltonian for closed loops $\Gamma$. The appropriate string and loop operator corresponds to \Eqref{eq:Sgamma} with $\alpha = 2\pi/3$:
\begin{equation}
S_\gamma^{\rm{QDPM}} \equiv S_\gamma \left( 2\pi/3\right), \quad W_{X,Y}^{\rm{QDPM}} \equiv W_{X,Y} \left( 2\pi/3\right). \label{eq:SQDPM}
\end{equation}
Thus we have 
\begin{equation}
\left[ W_{X,Y}^{\rm{QDPM}}, \HQDPM \right] = 0.
\end{equation}
and for closed curves $\Gamma$
\begin{equation}
\left[ S_\Gamma^{\rm{QDPM}}, \HQDPM \right] = 0.
\end{equation}

Our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of $\HQDPM$ are ergodic in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{\rm{QDPM}},W_Y^{\rm{QDPM}} \right).
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 0,1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X,Y}^{\rm{QDPM}}$ operators have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

We have demonstrated that ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \ref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a critically correlated or confining symmetry broken phase, which we discuss below.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$. The histograms of these U(1) winding numbers is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$. This confirms that $\Phi_X$ and $\Phi_Y$ are only conserved modulo $3$, rather than strictly conserved as they are in the QDM. We find that the histogram $P(\Phi)$ has a roughly Gaussian envelope with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{\rm{QDPM}},W_Y^{\rm{QDPM}})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in for the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

If a defect is introduced which violates the local constraint at a vertex, this can be viewed as an ``electric charge'' living on the vertex. One may determine if such electric charges are confined by introducing a pair of such defects and computing their spatial correlation function; if the free energy cost of defects at large distances remains finite, electric charges are deconfined in the GQDM ground state.

@@

We now relax the local constraint to a parity constraint such that only $Q_v^{Z_2} \equiv (Q_v \mod 2)$ is required to vanish; the resulting GQDM are the even and odd sectors are the  even \& odd Ising gauge theories (IGT) with constraints $n_v \in \{0,2,4\}$ and $n_v \in \{1,3\}$, respectively. In these cases, the local constraint requires $\alpha_v \in \{ 0, \pi\}$ in \Eqref{eq:Gv} and therefore the gauge symmetry is reduced from $U(1)$ to $Z_2$. The set of allowed dimer resonances $\mathcal{C}_{\rm{IGT}}$ is expanded accordingly:
\begin{align}
\mathcal{C}_{\rm{IGT}} &\equiv \bigl\{ \left(c_{\Box,i},\bar{c}_{\Box,i}\right) \bigr\} \\
&=\bigl\{
\left(\vrtplqt,\hrzplqt\right),\left(\threeleftplqt,\oneleftplqt\right),\left(\twodownplqt,\twoupplqt\right),\left(\bplqt,\fplqt\right) \bigr\}
\end{align}
where we implicitly include symmetry related resonances. We will limit or discussion to $v_\bplqt=0$ such that the corresponding Hamiltonian is
\begin{equation}
\HIGT\left(t\right) \equiv \HGQDM \biggl( \mathcal{C}_{\rm{IGT}},\bigl\{ \left(t,0\right) \bigr\} \biggr).
\end{equation}
In this case the equal superposition RK wavefunction is the exact ground state occurs in the absence of a potential energy term, as we can define a projection operator on each plaquette
\begin{equation}
h_\Box \equiv \frac{1}{2} \left( \mathbb{1} - \sum_i \eket{c_{\Box,i}}\ebra{\bar{c}_{\Box,i}} + h.c. \right)
\end{equation}
such that $\HIGT$ takes the form of \Eqref{eq:HRK} (up to a constant), and the RK state is the ground state for all values of $t$.

The increased dynamics of $\HIGT$ vs. $\HQDM$ no longer conserve the dimer flux $\Phi_\Gamma$ through a closed curve; only $\Phi_\Gamma \mod 2$ is preserved by $\HIGT$. Consequently, to define the string and loop operators that commute with $\HIGT$ (for closed strings) we fix $\alpha = \pi$ in \Eqref{eq:Sgamma}:
\begin{equation}
S_\gamma^{Z_2} \equiv S_\gamma \left( \pi \right), \quad W_{X,Y}^{Z_2} \equiv W_{X,Y} \left( \pi \right).
\end{equation}
Thus the $W_{X,Y}^{Z_2}$ have two eigenvalues $\pm 1$, and the topological sectors on the torus can be labeled as~$\Omega= (\pm 1,\pm1)$; consequently the extensive degeneracy of the fixed number constrained models has been reduced to a finite ground state degeneracy in the fixed parity models.

@@

We may understand the emergence of the $Z_2$ topologically ordered phase of IGTs from the QDM and QFPLM from the following picture. The additional vertex configurations that are allowed in the IGTs relative to the models with a dimer number constraint  have charge U(1) charge $Q_v = \pm 2$ (see \tabref{tab:defects}). It is known that coupling a $U(1)$ gauge field to a dynamics charge $q$ matter field can result in an $Z_q$ gauge theory~\cite{Fradkin1979}; as in the IGT's we have added dynamic charge 2 defects to the underlying $U(1)$ symmetric model, and we may expect a $Z_2$ gauge theory to result. We note that pentamers are charge $\pm3$ objects in the QDM (see \tabref{tab:defects}); thus below we investigate whether introducing pentamers into the QDM leads to an effective $Z_3$ gauge theory.

@@

The Hamiltonian of the QDPM is defined by \Eqref{eq:HGQDM} including both plaquette flips as well as local dimer resonances that create and destroy pairs of pentamers, and allow single pentamers to translate on the same sub-lattice; these local dynamics involving pentamers are illustrated in  \tabref{tab:pentamer_moves}. Note that pentamers give dynamics to ``staggered" regions with no flippable plaquettes; thus we expect the QDPM to have weaker correlations that the QDM, where the dynamics are more constrained. We may write $\HQDPM$ as the following:
\begin{equation}
\HQDPM \equiv \HQDM + H_{\rm{pen}}
\end{equation}
where $H_{\rm{pen}}$ only involves the pentamer resonances in \tabref{tab:pentamer_moves} and those related by symmetry:
\begin{align}
H_{\rm{pen}} \equiv \HGQDM \biggl( &\bigl\{ \left(c_{\nu_1},\bar{c}_{\nu_1}\right),\left(c_{\nu_2},\bar{c}_{\nu_2}\right),\left(c_{\nu_3},\bar{c}_{\nu_3}\right) \bigr\},\biggr.\notag\\
\biggl. &\bigl\{ \left(t_1,v_1\right),\left(t_2,v_2\right),\left(t_3,v_3\right) \bigr\} \biggr)  \label{eq:Hpen}
\end{align}

\begin{table}
    \begin{tabular}{| l | c | c |}
  	\hline
    $\nu$ & $c_\nu$ & $\bar{c}_\nu$ \\
	\hline
    $\nu_1$ & $\pentmvcreate$ &  $\pentmvanal$  \\
	\hline
    $\nu_2$ &$\penmvhrzleft$&$\penmvhrzright$ \\
	\hline
    $\nu_3$ &$\penmvdiagup$& $\penmvdiagdown$\\
	\hline
  \end{tabular}
  \caption{The pairs of local dimer \& pentamer configurations $c_\nu\leftrightarrow\bar{c}_\nu$ generate minimal local dynamics that create/annihilate pairs ($\nu_1$) of pentamers and translate single pentamers ($\nu_2$ \& $\nu_3$). Note that pentamers always stay on the same sublattice which is associated with their $U(1)$ charge. }
  \label{tab:pentamer_moves}
\end{table}

Because of the terms involving pentamers, $\HQDPM$ conserves $Q_v^{Z_3}$ rather than $Q_v$, and thus $\Phi_\Gamma$ does not commute with the Hamiltonian for closed loops $\Gamma$. The appropriate string and loop operator corresponds to \Eqref{eq:Sgamma} with $\alpha = 2\pi/3$:
\begin{equation}
S_\gamma^{Z_3} \equiv S_\gamma \left( 2\pi/3\right), \quad W_{X}^{Z_3} \equiv W_{X} \left( 2\pi/3\right) \label{eq:SQDPM}
\end{equation}
and similarly for $W_Y^{Z_3}$. With these definitions, closed string operators $S_\Gamma^{Z_3}$ and the topological loop operators $W_{X}^{Z_3}$ and $W_{Y}^{Z_3}$ commute with $\HQDPM$.

Our numerical simulations (discussed below in \secref{sec:numerics}) suggest that the dynamics of $\HQDPM$ are ergodic in each topological sector $\Omega$ that are labeled by the topological loop operators:
\begin{equation}
\Omega = \left(W_X^{Z_3},W_Y^{Z_3} \right).
\end{equation}
The exact zero energy ground state at the RK point, where $t_i=v_i, i \in \{ 1,2,3\}$, is the RK state, where the superposition is taken to be over all states in the topological sector. Since the $W_{X}^{Z_3}$ and  $W_{Y}^{Z_3}$ have eigenvalues $\{1,e^{i2\pi/3},e^{-i2\pi/3}\}$, the RK point of the QDPM is therefore 9-fold degenerate on the torus.

@@

We have demonstrated that ground state at the RK point of the QDPM has an exact $Z_3$ gauge symmetry and a corresponding $9$-fold topological degeneracy on the torus. We emphasize that despite the exact $Z_3$ local gauge symmetry of the QDPM, there is not an exact mapping to a $Z_3$ gauge theory on the square lattice, as the QDPM has Ising degrees of freedom on the links, rather than the 3 dimensional local Hilbert space of a $Z_3$ gauge theory (see \ref{sec:Z3GT} for details of $Z_3$ gauge theory).  It remains to address whether this RK state is in the deconfined $Z_3$ topologically ordered phase of a $Z_3$ gauge theory, or a confined symmetry broken or critical state, which we discuss below.

@@

We explicitly demonstrate how the extensive topological degeneracy of the QDM is reduced to a finite degeneracy in the QDPM by computing the histogram of $U(1)$ topological winding numbers $\Phi_{X}$ and $\Phi_{Y}$. The histograms of these U(1) winding numbers is shown in \figref{fig:u1_wind_qdpm}, where the different colors represent calculations that were initialized to different winding sectors $(\Phi_X,\Phi_Y)$. This confirms that $\Phi_X$ and $\Phi_Y$ are only conserved modulo $3$, rather than strictly conserved as they are in the QDM. We find that the histogram $P(\Phi)$ has a Gaussian shape with a width that increases with system size; this is expected as $\Phi = \pm L/2$ correspond to ordered configurations and thus there are fewer dimer configurations near these limits. We find no evidence of a lack of ergodicity within each of the 9 topological sectors. For the remainder of the paper we present results in the $(0,0)$ topological sector, but we see no qualitative dependence between topological sectors for any of the results presented here.  Thus we find that the QDPM has a \emph{finite} $9$-fold degeneracy on the torus, where the topological sectors are labeled $(W_X^{Z_3},W_Y^{Z_3})$, in contrast with the extensive degeneracy of the QDM.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{u1_wind_qdpm.pdf}
    \caption{The histogram of the $U(1)$ topological winding numbers $\Phi_X$ and $\Phi_Y$ as measured in for the ground state of the RK point of the QDPM for three different lattice sizes. The colors correspond to calculations initialized to different topological sectors $(\Phi_X,\Phi_Y)$: blue $(0,0)$, red $(1,1)$, and purple $(-1,-1)$. The lines are fits to a Gaussian for each system size. Note that $\Phi$ is conserved modulo $3$, resulting in a $9$-fold topological degeneracy on a torus.}
    \label{fig:u1_wind_qdpm}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
